# Data Cleaning and Reformatting Variables

In this section, we consider how to convert the list of character matrices, `men_res_mat`, into an appropriate format for analysis. Currently, the data values are all character, which is not conducive to, e.g., finding the median age of the runners. However, we can easily reformat age into numeric values with the `as.numeric()` function. Do we want to turn the entire matrix into a numeric matrix? Not really. It doesn’t make sense to try to convert the runner’s name into a numeric value. For this reason, we want to create a data frame because it allows our variables to be different types. We have 6 variables: the runner’s name, home town, age, and 3 versions of time. As just mentioned, we want to convert age to a numeric and leave name as character. What about the other variables? We probably want to also keep hometown as character.

Time is stored as a string in the format: `hh:mm:ss`. We want time in a numeric format so it can be more easily summarized and modeled. For this, we will use the `lubridate` package, which is excellent for converting character strings to dates and times. We also need to reconcile the 3 different recorded times (gun, net, and plain time). Net time is considered more accurate than gun time so we can simply use net time when available and otherwise use gun time or time, whichever is reported. Of course, we can keep all 3 versions of time around and let the analyst explore relationships between them and decide which to use, but we keep things simple for now and just report one time for each runner.

Before we begin converting our character strings into numeric values, we also consider whether there are any new variables we might want to create. If we are to combine all the data from the 14 years of records into one data frame, then we should keep track of the year. Likewise, if we are to combine the men’s and women’s results then we also want a variable that indicates the sex of the runner. These are very simple to make when combining data frames with `dplyr::bind_rows()`.

We begin with the task of creating the numeric variable `age` with `as.numeric()`, e.g., for the 2012 males,

```{r}
age <- as.numeric(men_res_mat$`2012`[ ,'ag'])
```

Note that we subsetted the list to work with the 2012 matrix and then subsetted this matrix to work with the column named `ag`. We check a few `age` values with

```{r}
tail(age)
```

These values look reasonable, but let’s check more thoroughly that our data extraction works as expected by summarizing each year’s ages with

```{r, warning=T}
age <- map(men_res_mat, ~ as.numeric(.x[ ,'ag']))
```

We received a warning message that our conversion of the character values for age into numeric resulted in `NA` values, meaning that some of the values do not correspond to numbers. We want to look into the specific cause of these messages, but first we examine `age`.

We create side-by-side boxplots of the yearly distribution of the age of the runners to give a quick check on the reasonableness of the values.

```{r ageboxplot, fig.cap="Box Plot of Age by Year. These side-by-side boxplots of age for each race year show a few problems with the data for 2003 and 2006. The runners in these years are unusually young."}
age %>% 
  enframe(name = 'year', value = 'age') %>% 
  unnest() %>% 
  ggplot(aes(year, age)) +
  geom_boxplot()
```

Figure \@ref(fig:ageboxplot) reveals problems with 2 years. All of the runners in 2003 were under 10 and more than 1 in 4 runners in 2006 were under 10! Clearly something has gone wrong.

Let's examine the original text for 2003 and 2006.

```{r}
head(men_files$`2003`)
```

```{r}
men_files$`2006`[2200:2205]
```

We see that in 2003, the age values are shifted to the right one space in comparison to the location of the ‘=’ characters. This means that we are picking up only the digit in the tens place. In 2006, some but not all of the rows have values that are off by one character.

We can easily solve both of these problems by including the value in the “blank” space between columns. We can do this by changing the index for the end of each variable when we perform the extraction. That is, we modify the line in `select_cols()` that locates the end of a column to include the blank position, i.e.,

```{r}
select_cols <- function(col_names, header_row, search_locs) {
  select_col <- function(name, header_row, search_locs) {
    start_pos <- str_locate(header_row, name)[1,1]
    if (is.na(start_pos)) return(c(NA, NA))
    index <- sum(start_pos >= search_locs)
    c(search_locs[index] + 1, search_locs[index + 1])
  }
  
  map(col_names, select_col, 
      header_row = header_row, 
      search_locs = search_locs) %>% 
    do.call('rbind', .)
}
```

When we use this revised calculation in `select_cols()`, we pick up the blank character after each field. This should not matter when we convert our text data to numeric and if we don’t want trailing blanks in our character-valued variables, we can easily remove them with regular expressions.

```{r include=FALSE}
men_res_mat <- map(men_files, extract_variables)
women_res_mat <- map(women_files, extract_variables)
age <- map(men_res_mat, ~ as.numeric(.x[ ,'ag']))
```

In the process of confirming our conversion of age from character to numeric, we uncovered problems with our extraction process. We need to modify our helper function `select_cols()` from Chapter \@ref(reading) to address the problem. This process is iterative as we continue to check that our data make sense. When we uncover nonsensical results, we investigate them further, which possibly leads to retracing our steps to clean up messy data.

After we modify this one line of code in `select_cols()` and reapply this updated version of the function to the tables of race results, we check again the summary statistics with boxplots. We find that the problem with too many young runners has cleared up (see Figure \@ref(fig:ageboxplots-fixed)).

```{r ageboxplots-fixed, echo=F, fig.cap="Box Plot of Age by Year. These side-by-side boxplots of age for each race year show a reasonable age distribution. For example, the lower quartile for all years range between 29 and 32. The problems identified earlier for 2003 and 2006 have been addressed."}
age %>% 
  enframe(name = 'year', value = 'age') %>% 
  unnest() %>% 
  ggplot(aes(year, age)) +
  geom_boxplot()
```


We now turn to the warning messages that occurred when we converted the character strings for age to numeric values. We were given several messages ‘NAs introduced by coercion’. We count the number of `NA` values in each year with

```{r}
map_int(age, ~ sum(is.na(.)))
```

In 2001, we have 62 `NA`s for age. We need to investigate. To make our work simpler, let's assign the 2001 ages to a vector called `age2001`. We do this with

```{r}
age2001 <- age$`2001`
```

Let's examine the original rows in the file that correspond to an `NA` in `age2001`. Recall that we dropped the header of the file before extracting the variables so we need to add an offset to the location of the `NA`s in `age2001` in order to pick out the correct rows in the original table. We find the offset with

```{r}
offset <- str_which(men_files$`2001`, '^===')
offset
```

We then find the lines in the original file that have the bad age values with

```{r}
bad_age_idx <- which(is.na(age2001)) + offset
tail(men_files$`2001`[bad_age_idx])
```

To save space, only the last few lines are shown here. With one exception, all of the rows are blank/empty. The one exception is the row that corresponds to the footnote that defines the meaning of the '#' annotation. Where in the table are these rows located?

```{r}
bad_age_idx
```

These blank lines are scattered throughout the file. We can modify the extraction by checking for blank rows and removing them. The regular expression,

```{r}
blanks <- str_which(men_files[['2001']], '^[[:blank:]]*$')
```

locates all rows that are entirely blank. The second argument to `str_which` uses several meta characters to specify the pattern to search for. The `^` is an anchor for denoting the start of the string, the `$` anchors to the end of the string, the `[[:blank:]]` denotes the equivalence class of space and tab characters, and the `*` indicates that the blank character can appear 0 or more times. All together the pattern `^[[:blank:]]*$` matches a string that contains any number of blanks from start to end, i.e., only blank lines.

A simpler expression locates the footnote rows, i.e., rows that begin with # or *. Here is the modified `extract_variables()` function:

```{r}
extract_variables <- function(file, var_names = c("name", "home", "ag", "gun", "net", "time")) {
  # find the index of the row with =s
  eq_idx <- str_which(file, '^===')
  # extract the two key rows and the data
  spacer_row <- file[eq_idx]
  header_row <- file[eq_idx - 1] %>% str_to_lower()
  # find blank lines
  blanks <- str_which(file, '^[[:blank:]]*$')
  # find comments
  comments <- str_which(file, '^[[:blank:]]*[#\\*]')
  # remove header, blank lines, and comments
  body <- file[-c(1:eq_idx, blanks, comments)]
  
  # get starting and ending positions of variables
  search_locs <- find_col_locs(spacer_row)
  loc_cols <- select_cols(var_names, header_row, search_locs)
  values <- map(body, str_sub, loc_cols) %>% do.call('rbind', .)
  colnames(values) <- var_names
  invisible(values)
}
```

```{r include=FALSE}
men_res_mat <- map(men_files, extract_variables)
women_res_mat <- map(women_files, extract_variables)
```

After adding this code to carry out the additional cleaning of the tables, the 61 `NA`s in 2001 are gone as well as many but not all of the other `NA`s in other the years.

```{r}
map_int(age, ~ sum(is.na(.)))
```

Continued inspection of Figure \@ref(fig:ageboxplots-fixed) uncovers another problem – the minimum values for age in 2001, 2002, and 2003 remain small, i.e., close to 0. That’s clearly not possible! Let’s find which runners have an age under 5 and look at their records in the original table. For 2001, we have

```{r}
which(age2001 < 5)
men_files[['2001']][which(age2001 < 5) + offset]
```

Apparently there are runners with an age entered as 0! Since these are the actual values in the table, we leave the decision as to what to do with these runners for later when we analyze the data. At this point, it appears we have successfully taken care of the creation of the age variable. However, we typically clean the variables simultaneously as an error in one variable often leads to errors in others based on position. As we clean the other variables, we may need to re-examine age to ensure that the values for age remain valid.

Next, we turn to the creation of the time variable. As mentioned at the beginning of this section, the time appears as `hh:mm:ss` and we wish to convert it to minutes. However, to carry out this computation, we must split the time field up into its constituent pieces. Also, some runners completed the race in under one hour so their times appear in a slightly different format, i.e., `mm:ss`, and we need to be able to handle both formats in our processing. For simplicity, we again start with converting the time variable for one year, say 2012. We create a vector to develop our code as follows:

```{r}
char_time <- men_res_mat[['2012']][ ,'time']
head(char_time)
tail(char_time)
```

We split each character string up into its parts using `str_split()` with

```{r}
time_pieces <- str_split(char_time, ':')
```

The `:` characters are discarded in the process, and the return value from `str_split()` is a list of character vectors. We have one vector for each input string, where the elements of the vector contain the pieces of the string separated by each `:` character. We confirm that the splitting worked properly by examining the first and last times, i.e.,

```{r}
head(time_pieces, 1)
tail(time_pieces, 1)
```

We convert these elements to numeric values and combine them into one value that reports time in minutes with

```{r}
run_time <- map_dbl(time_pieces, function(x) {
  x <- as.numeric(x)
  if (length(x) == 2) x[1] + x[2] / 60
  else 60 * x[1] + x[2] + x[3] / 60
})
```

We check our conversion with 

```{r}
summary(run_time)
```

It appears that our time conversion works. We saw earlier that the fastest runner completed the 2012 race in 45 minutes and 15 seconds, which is 45.25 minutes, and the slowest completed it in 2 hours 30 minutes and 59 seconds, which is nearly 151 minutes. For the sake of reproducibility, we encapsulate this conversion into a function called `convert_time()`.

```{r include=F}
convert_time <- function(t) {
  time_pieces <- str_split(t, ':')
  map_dbl(time_pieces, function(x) {
    x <- as.numeric(x)
    if (length(x) == 2) x[1] + x[2] / 60
    else 60 * x[1] + x[2] + x[3] / 60
  })
}
```


Let’s wrap these conversions into a function to apply to the character matrices in menResMat and return a data frame with variables for analysis. We call this function `create_df()`. In addition to the conversion of character strings to numeric, we also create two new variables, year and sex. To do this, we must have input arguments to tell us which year we are cleaning and whether the results are for men or women. Lastly, we also choose which time variable to include in the data frame from among the 3 available, with a preference for net time. The function appears as

```{r}
create_df <- function(res, year, sex) {
  # determine which time to use
  use_time <- if (!is.na(res[1,'net'])) {
                res[ ,'net']
              } else if (!is.na(res[1,'gun'])) {
                res[ ,'gun']
              } else {
                res[ ,'time']
              }
  run_time <- convert_time(use_time)
  
  results <- data_frame(year = year,
                        sex = sex,
                        name = res[ ,'name'],
                        home = res[ ,'home'],
                        age = as.numeric(res[ ,'ag']),
                        time = run_time)
  
  invisible(results)
}
```

We apply our new function, `create_df()`, to all of the male results as follows:

```{r message=FALSE, warning=FALSE}
men_df <- map2(men_res_mat, 1999:2012, create_df, sex = 'm')
warnings()[c(1, 2, 49, 50)]
```

It is likely that the conversion problems are coming from the conversion of time from a character string into minutes because we have already handled the conversion of age. We can check the number of `NA` values for `time` with

```{r}
map_dbl(men_df, ~ sum(is.na(.x$time)))
```

There are a large number of `NA`s in 2007, 2009, and 2010, and it appears that all of the run time values for 2006 are `NA`.

Let’s begin by examining a few of the records in 2007, 2009, and 2010 that have an `NA` in run time. We find that these are caused by runners who completed half the race but have no final times and by runners who have a footnote after their time, e.g.,

```{r, echo=F}
men_files[['2007']][c(8, 5280)]
```

We can easily modify `create_df()` to eliminate the footnote symbols (# and *) from the times and drop records of runners who do not complete the race. These revisions are

```{r}
create_df <- function(res, year, sex) {
  # determine which time to use
  use_time <- if (!is.na(res[1,'net'])) {
                res[ ,'net']
              } else if (!is.na(res[1,'gun'])) {
                res[ ,'gun']
              } else {
                res[ ,'time']
              }
  # remove #, *, and blanks from time
  use_time <- str_replace_all(use_time, '[#\\*[:blank:]]', '')
  # drop rows with no time
  res <- res[use_time != '', ]
  run_time <- convert_time(use_time[use_time != ''])
  
  results <- data_frame(year = year,
                        sex = sex,
                        name = res[ ,'name'],
                        home = res[ ,'home'],
                        age = as.numeric(res[ ,'ag']),
                        time = run_time)
  
  invisible(results)
}
```

After we apply this revised function to `men_res_mat` to create our data frame, most missing values are gone except for 2006.

```{r, include=F}
men_df <- map2(men_res_mat, 1999:2012, create_df, sex = 'm')
```

```{r}
map_dbl(men_df, ~ sum(is.na(.x$time)))
```

Let's look at the missing value for 2001.

```{r}
men_df[['2001']] %>% filter(is.na(time))
i <- grep('Peter HUI', men_files[['2001']])
men_files[['2001']][(i-1):(i+1)]
```

Somehow this runner's time got entered incorrectly. Because runners are listed in order of their time, we can infer that he probably had a time of 1:31:47. We'll fill in that value manually.

```{r}
men_df[['2001']]$time[2250] <- 91 + 47/60
```

As for 2006, the problem can be seen when we look at the header lines of that file.

```{r}
men_files[['2006']][7:11]
```

The separator line of '='s does not have a space between the "Hometown" and "Net Tim" columns. The easiest thing to do in this scenario is to just alter the separator line directly and rerun `extract_variables()`.

```{r}
spacer_row_2006 <- men_files[['2006']][8]
str_sub(spacer_row_2006, 64, 64) <- ' '
men_files[['2006']][8] <- spacer_row_2006
men_files[['2006']][7:11]
```

At last, we combine the race results for all years and men into one data frame using `purrr::map2_dfr()`.

```{r, include=F}
men_res_mat <- map(men_files, extract_variables)
men_df <- map2_dfr(men_res_mat, 1999:2012, create_df, sex = 'm') %>% 
  mutate_if(is_character, funs(str_trim(., side = 'both'))) %>% 
  mutate(name = str_to_title(name))
men_df[which(men_df$year == 2001 & men_df$name == 'Peter Hui'),'time'] <- 91 + 47/60
```

```{r}
write_csv(men_df, 'men/men_df.csv')
```

We check the dimension of our amalgamated data frame with

```{r}
dim(men_df)
```

We also examine a summary of the variables in cbMen to check whether any problems arose, e.g., with coercion, in the binding together of the data frames.

Over these 14 years, 70,070 male runners completed the Cherry Blossom race. In addition, more than 70,000 female runners completed the race. We leave it as an exercise to handle the women’s race results. In the next section we take a closer look at the race results.

```{r message=FALSE, warning=FALSE}
women_res_mat <- map(women_files, extract_variables)
women_files[['2006']][8] <- spacer_row_2006
women_df <- map2_dfr(women_res_mat, 1999:2012, create_df, sex = 'f')
women_df[grep('^Marie-Laure Poir', women_df$name),'name'] <- 'Marie-Laure Poire'
women_df <- women_df %>%
  mutate_if(is_character, funs(str_trim(., side = 'both'))) %>% 
  mutate(name = str_to_title(name))
write_csv(women_df, 'women/women_df.csv')
```

